---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MaddisonData

<!-- badges: start -->
[![`R-CMD-check`](https://github.com/sbgraves237/MaddisonData/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sbgraves237/MaddisonData/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Make it easier for humans to access data from the Maddison Data Project in R. 
Later releases may include vignettes, etc., documenting analyses using the 
[`KFAS`] (Kalman filtering and smoothing, aka state space) techniques with 
these data.  

Objectives: Make it relatively easy in R to do the following: 

1. Find the countries with the highest `gdppc` for each year for which data are 
available using `MaddisonLeaders()`. 

2. Refine "1" by deleting companies with high `gdppc` based on something narrow 
like a commodity, e.g., oil, since 1600: 

```{r MadDat1600}
library(MaddisonData)
MadDat1600 <- subset(MaddisonData, year>1600)
Leaders1600 <- MaddisonLeaders(c('ARE', 'KWT', 'QAT'), data=MadDat1600)
summary(Leaders1600)
```

3. Plot the data available on `gdppc` and / or pop for a selection of countries, 
e.g., world leaders. 

```{r ggplotPath}
str(GBR_USA <- subset(MaddisonData::MaddisonData, ISO %in% c('GBR', 'USA')))
GBR_USA1 <- MaddisonData::ggplotPath('year', 'gdppc', 'ISO', GBR_USA, 1000)

GBR_USA1+ggplot2::coord_cartesian(xlim=c(1500, 1850)) # for only 1500-1850 
GBR_USA1+ggplot2::coord_cartesian(xlim=c(1600, 1700), ylim=c(7, 17)) 

# label the lines
ISOll <- data.frame(x=c(1500, 1800), y=c(2.5, 1.7), label=c('GBR', 'USA'), 
              srt=c(0, 30), col=c('red', 'green'), size=c(2, 9))
GBR_USA2 <- ggplotPath('year', 'gdppc', 'ISO', GBR_USA, 1000, 
                    labels=ISOll, fontsize = 20)  
                        
# h, vlines, manual legend only 
Hlines <- c(1,3, 10, 30)
Vlines = c(1849, 1929, 1933, 1939, 1945)
(GBR_USA3 <- ggplotPath('year', 'gdppc', 'ISO', GBR_USA, 1000, 
       ylab='GDP per capita (2011 PPP K$)', 
       legend.position = NULL, hlines=Hlines, vlines=Vlines, labels=ISOll))  
```

LATER: 

4. Build a state space / Kalman models for `gdppc` and `pop` for each country 
in the Maddison project data. 

5. Use Kalman smooth to interpolate and extrapolate (forward but not 
backwards) `gdppc` and `pop` for each country for all years that appear 
anywhere in the Maddison project data. 

6. Identify the world leader in `gdppc` for each year, refining "1" using 
`KFAS` interpolation. 

7. Identify the world technology leader for each year by evaluating the 
`gdppc` leader for each year and replacing any whose leadership was narrow 
like members of OPEC with a country with a broad-based economy like the US. 

## Installation

You can install the development version of MaddisonData from 
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("sbgraves237/MaddisonData")
```

## Example

[Coming soon.]
<!--This is a basic example which shows you how to solve a common problem:-->

```{r example}
library(MaddisonData)
## basic example code
```